Handling String-Number Conversion in String Constraint Solving

### 2019-11-13 draft.tex
polish 5.2 and 5.3

For the purpose of program analysis, string solver also needs to support \emph{equality constraints} (e.g. $x.y = y.x$), \emph{membership constraints} (e.g., $x \in (a.b)^*$), and \emph{integer constraints} (e.g., $|x|-|y|>3$). Most of the standard string operations can be encoded using the above three types of constraints. For instance, $y=x.charAt(i)$ can be encoded as $x=x_1.x_2.x_3 \wedge |x_1| = i \wedge |x_2| =1 \wedge y= x_2$. Those are also the type of constraints supported by most state-of-the-art string constraint solvers such as Z3, CVC4, Z3Str3, and Trau+.

@@

Our string constraint solving produce can be easily integrated in a \emph{satisfiability module theory (SMT)} solver as a theory solver. The integration allows us to express string operations that cannot be reduced to the type of constraints (membership, equality, integer, string-number conversion) supported by our procedure. For example, we can encode \texttt{split} using the code fragment in Figure~\ref{fig:smt_string}. Other functions such as \texttt{replaceAll} and \texttt{reverse} can be encoded in a similar manner. The performance of such encoding might not be as good as those that provide direct support to those operations, but at least it works for some cases in our experiments. Further details of the combination as well as some other implementation details can be found in Section~\ref{section:implementation}.

@@

\hide{
To ease presentation, in the paper, we only consider string constraints consisting of \emph{equality constraints (a.k.a. word equation)}, \emph{membership constraints}, \emph{length constraints}, and \emph{string-number conversion} functions.

@@

A \emph{membership constraint} $\phi_m$ is of the form $x \in L(A)$ where $x$ is a string variable and $A$ is a finite automaton. The \emph{model} of $\phi_m$ is the set of interpretations $\modelsof{\phi_m}=\{I\mid 
I(x) \in L(A) \}$.

@@

More concretely, at the beginning, we move all integer constraints in $\Phi_I$ to the output linear formulas $\Phi_O$. After this step, $\Phi_I$ is the union of the three sets of constraints: the equality constraints, the membership constraints, and the string-number conversion constraints. For all string variables $x$ that occurs in a string-number conversion constraint $n= \sti{x}$, we follow the rule that will be described in Section~\ref{section:s2i} to pick the corresponding PFA. We do not set any restriction to the selection of PFA for other string variables. The selection is an implementation decision. We will describe the PFA selection strategy of \tool in Section~\ref{section:implementation}.

After the PFA for each variable is selected, we step-wisely pick and remove a string constraint from $\Phi_I$, executing the corresponding procedure to create equisatisfiable linear constraints, and add all the results to $\Phi_O$. The procedure for equality and membership constraints is in Section~\ref{section:eq} and the one for string-number conversion is in Section~\ref{section:s2i}.

@@

\yfc{say handling membership constraint can be exponential to the number of constraints, but in our under-approximation, it is polynomial.}

@@

%We actually have even something stronger, namely, that, $\syncfof A {A'}$ has all possible allignments. How do I say this ... or do I have to say this?
%\begin{lemma}
%For $\pim:\pvarone\rightarrow \nat$, $\pim':\pvartwo \rightarrow\nat$, $I:\cvarone\rightarrow\Sigma$, $I':\cvartwo\rightarrow\Sigma$, 
%we have $(I\cup\pim)\cup(I'\cup\pim')\in\semof {\syncfof A {A'}}$ if and only if $\decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')$.
%\end{lemma}

@@

\subsection{Handling Equalities and Membership Constraints Symbolically} \label{section:eq}

Assume that the constraint we take from $\Phi_I$ is an equality constraint $\phi_e= x_1\cdot x_2 \cdots x_n = x_{n+1}\cdot x_{n+2} \cdots x_m$, where each string variable $x_i$ is restricted to a PFA $A_{i}$ with the set of character variables $V_{i}$, for $i\in[1,m]$. We first build two PFAs that encode the left-hand side term and the right-hand side term of the equality constraint. Intuitively, this can be achieved by connecting the PFAs of the string variables with epsilon transitions. However, there are two minor technical issues. First, our definition of PFA does not allow epsilon transition. Second, some string variable might occur multiple times in a equality constraint, but the definition of PFA allows each character variable occurs in only one transition.

We circumvent the first problem by introducing some fresh characters variables, say $v_{\epsilon_n}$, and forcing its value to be $\epsilon$ by adding new linear formulas, say $v_{\epsilon_n}= -1$, to the output linear formulas $\phi_O$. Then we can connect the final state and initial state of two consecutive variables with an ``epsilon'' transition, i.e., one with the label $v_{\epsilon_n}$. We use $V_\epsilon$ to denote the set of all variables used for ``epsilon'' transitions.

@@

Now let $A$ and $A'$ over character variables $V$ and $V'$ be the two PFAs obtained from the two sides of the equality constraint $\phi_e$. We then construct the synchronization formula $\syncfof A {A'}$ and add it to $\phi_O$. Recall that from Lemma~\ref{lemma:synccorrect}, each model of $\syncfof A {A'}$ corresponds to a pair of words in $\semof{A}$ and $\semof{A'}$. However, this does not directly correspond to a model of the equality constraint $\phi_e$. The reason is that difference occurrence of the same string variables can be assigned different values in $\semof{A}$ and $\semof{A'}$. To avoid this problem, we need to add more constraints to ensure that all variables renamed from the same original variable are assigned the same value and are occurred equally often.

$$ \Psi_r \defeq
\bigwedge_{v\in V_x, v'\in R_v} v=v' \wedge \#v = \#v'
$$
Here $V_x$ is the set of all character variables occurred in the PFAs of $x_1,x_2,\ldots,x_m$, i.e., $V_x=\bigcup_{i\in [1,m]}{V_{i}}$.

@@

\begin{lemma}\label{lemma:eqcorrect}
	%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
	$\semof {\syncfof A {A'}  \wedge \Psi_r}_{\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo \setminus V_{\epsilon}} = \{\bigcup_{i\in[1,m]}(I^i\cup\pim^i)\mid w_1\cdot w_2 \cdots w_n = w_{n+1}\cdot w_{n+2} \cdots w_m \wedge \bigwedge_{i\in [1,m]}w_i=\decode A_{i}(I^i\cup\pim^i) \}$. 
\end{lemma}

>>>>

For the purpose of program analysis, string solver also needs to support \emph{equality constraints} (e.g. $x.y = y.x$), \emph{regular constraints} (e.g., $x \in (a.b)^*$), and \emph{integer constraints} (e.g., $|x|-|y|>3$). Most of the standard string operations can be encoded using the above three types of constraints. For instance, $y=x.charAt(i)$ can be encoded as $x=x_1.x_2.x_3 \wedge |x_1| = i \wedge |x_2| =1 \wedge y= x_2$. Those are also the type of constraints supported by most state-of-the-art string constraint solvers such as Z3, CVC4, Z3Str3, and Trau+.

@@

Our string constraint solving produce can be easily integrated in a \emph{satisfiability module theory (SMT)} solver as a theory solver. The integration allows us to express string operations that cannot be reduced to the type of constraints (regular, equality, integer, string-number conversion) supported by our procedure. For example, we can encode \texttt{split} using the code fragment in Figure~\ref{fig:smt_string}. Other functions such as \texttt{replaceAll} and \texttt{reverse} can be encoded in a similar manner. The performance of such encoding might not be as good as those that provide direct support to those operations, but at least it works for some cases in our experiments. Further details of the combination as well as some other implementation details can be found in Section~\ref{section:implementation}.

@@

\hide{
To ease presentation, in the paper, we only consider string constraints consisting of \emph{equality constraints (a.k.a. word equation)}, \emph{regular constraints}, \emph{length constraints}, and \emph{string-number conversion} functions.

@@

A \emph{regular constraint} $\phi_r$ is of the form $x \in L(A)$ where $x$ is a string variable and $A$ is a finite automaton. The \emph{model} of $\phi_m$ is the set of interpretations $\modelsof{\phi_m}=\{I\mid 
I(x) \in L(A) \}$.

@@

More concretely, at the beginning, we move all integer constraints in $\Phi_I$ to the output linear formulas $\Phi_O$. After this step, $\Phi_I$ is the union of the three sets of constraints: the equality constraints, the regular constraints, and the string-number conversion constraints. For all string variables $x$ that occurs in a string-number conversion constraint $n= \sti{x}$, we follow the rule that will be described in Section~\ref{section:s2i} to pick the corresponding PFA. We do not set any restriction to the selection of PFA for other string variables. The selection is an implementation decision. We will describe the PFA selection strategy of \tool in Section~\ref{section:implementation}.

After the PFA for each variable is selected, we step-wisely pick and remove a string constraint from $\Phi_I$, executing the corresponding procedure to create equisatisfiable linear constraints, and add all the results to $\Phi_O$. The procedures for regular, equality and string-number conversion constraints are in Sections~\ref{section:mem},~\ref{section:eq}, and~\ref{section:s2i}, respectively.

@@

\yfc{say handling regular constraint can be exponential to the number of constraints, but in our under-approximation, it is polynomial.}

@@

%We actually have even something stronger, namely, that, $\syncfof A {A'}$ has all possible allignments. How do I say this ... or do I have to say this?
%\begin{lemma}
%For $\pim:\pvarone\rightarrow \nat$, $\pim':\pvartwo \rightarrow\nat$, $I:\cvarone\rightarrow\Sigma$, $I':\cvartwo\rightarrow\Sigma$, 
%we have $(I\cup\pim)\cup(I'\cup\pim')\in\semof {\syncfof A {A'}}$ if and only if $\decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')$.
%\end{lemma}
In the rest of the section, we assume each string variable $x$ is restricted to a PFA $A_x$ with the set of character variables~$V_x$.

@@

\subsection{Handling Regular Constraints Symbolically} \label{section:mem}
Assume that the constraint we take from $\Phi_I$ is a regular constraint $\phi_r \defeq x \in L(A)$. Assume that $A=(Q,T,\Sigma,q_i,q_f)$. First we create a PFA $A' = (Q,T', V',q_i,q_f)$ that has exactly the same set of states as $A$. The set of transitions $T'$ and character variables $V'$ can be obtained by the following procedure.
Initially, both $V'$ and $T'$ are empty sets.
For each transition $(q, a, q') \in T$, we add a fresh character variable $v$ to $V'$, add the transition $(q,v,q')$ to $T'$, and add the constraint $v = \modelsof{a}$ to $\Phi_O$. Recall that $\modelsof{a}$ is the numerical encoding of the character $a$. After $A'$ is built, we add the synchronization formula $\syncfof {A_x }{A'}$ to $\Phi_O$.

The correctness of this construction is expressed in Lemma~\ref{lemma:memcorrect} below. 
%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}
\begin{lemma}\label{lemma:memcorrect}
	%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
	${\semof {\syncfof {A_x }{A'}}}_{V_x \cup \#V_x} = \{(I\cup\pim) \mid \decode {A_x}(I\cup\pim) \in L(A)\}$. 
\end{lemma}

\subsection{Handling Equality Constraints Symbolically} \label{section:eq}

Assume that the constraint we take from $\Phi_I$ is an equality constraint $\phi_e\defeq x_1\cdot x_2 \cdots x_n = x_{n+1}\cdot x_{n+2} \cdots x_m$. We first build two PFAs that encode the left-hand side term and the right-hand side term of the equality constraint. Intuitively, this can be achieved by connecting the PFAs of the string variables with epsilon transitions. However, there are two minor technical issues. First, our definition of PFA does not allow epsilon transition. Second, some string variable might occur multiple times in a equality constraint, but the definition of PFA allows each character variable occurs in only one transition.

We circumvent the first problem by introducing some fresh characters variables, say $v_{\epsilon_n}$, and forcing its value to be $\epsilon$ by adding new linear formulas, say $v_{\epsilon_n}= -1$, to the output linear formulas $\phi_O$. Then we can connect the final state and initial state of two consecutive variables with an ``epsilon'' transition, i.e., one with the label $v_{\epsilon_n}$. %We use $V_\epsilon$ to denote the set of all variables used for ``epsilon'' transitions.

@@

Now let $A$ and $A'$ over character variables $V$ and $V'$ be the two PFAs obtained from the two sides of the equality constraint $\phi_e$. We then construct the synchronization formula $\syncfof A {A'}$ and add it to $\Phi_O$. Recall that from Lemma~\ref{lemma:synccorrect}, each model of $\syncfof A {A'}$ corresponds to a pair of words in $\semof{A}$ and $\semof{A'}$. However, this does not directly correspond to a model of the equality constraint $\phi_e$. The reason is that difference occurrence of the same string variables can be assigned different values in $\semof{A}$ and $\semof{A'}$. To avoid this problem, we need to add the constraint $\Psi_r$ below to $\Phi_O$ to ensure that all variables renamed from the same original variable are assigned the same value and are occurred equally often.

$$ \Psi_r \defeq
\bigwedge_{v\in V_X, v'\in R_v} v=v' \wedge \#v = \#v'
$$
Here $V_X$ is the set of all character variables occurred in the PFAs of $x_1,x_2,\ldots,x_m$, i.e., $V_X=\bigcup_{i\in [1,m]}{V_{x_i}}$.

@@

\begin{lemma}\label{lemma:eqcorrect}
	$\semof {\syncfof A {A'}  \wedge \Psi_r}_{V_X \cup \#V_X} = \{\bigcup_{i\in[1,m]}(I^i\cup\pim^i)\mid w_1\cdot w_2 \cdots w_n = w_{n+1}\cdot w_{n+2} \cdots w_m \wedge \bigwedge_{i\in [1,m]}w_i=\decode A_{x_i}(I^i\cup\pim^i) \}$. 
\end{lemma}
